# Data generating changes

```{r}
#| label: setup
#| message: false
#| warning: false

source("./setup-and-data-prep.R")

n_zero <- dta_for_datacheck |>
  filter(tor_length == 0 | length_mi== 0) |>
  nrow()

```

As noted in the introduction, the data-generating processes have changed since 1970: over time the records include less missing data, more short-lived weak tornadoes, and more accurate coordinates.

## Changes in tornado detection over time {#sec-changes-detection-over-time}

The plots below reveal that prior to 1997 weak tornadoes and ones with short tracks were mostly missed or the data was incomplete. I presume the improvements were are due to the introduction of more automated means of identifying and estimating the force of these less significant tornadoes, for example, directly from radar or satellite data with sensors providing higher resolution data, and NWS staff using more powerful computers and more sophisticated algorithms (which seem more likely than a change in the frequency of weak, short-lived tornadoes over this time period).

@fig-changes-in-detection-over-time provides evidence that the NWS offices started using better instruments. The reduction in missing coordinate data starting in 1996 is very impressive (panel A). So is the increase in the number of records of short-lived, weak tornadoes evident in panels B and C starting in the 2000s.

```{r, fig.height=8, fig.width=12}
#| label: fig-changes-in-detection-over-time
#| fig-cap: "Are there changes in detection over time?"
#| fig-height: 8
#| fig-width: 12
#| column: page-inset-right

p1 <- dta_non_mapping |>
  mutate(missing_data = is.na(begin_lat) | is.na(end_lat)) |> # if lat is missing, lon is missing too
  summarize(n = n(),
            n_missing = sum(missing_data),
            n_not_missing = sum(!missing_data),
            pct_missing = n_missing / n,
            .by = year) |>
  ggplot() +
  geom_jitter(aes(year, pct_missing, size = n),
              alpha = 0.6) + #size = 1, 
  scale_y_continuous(labels = label_percent(),
                     expand = expansion(mult = c(0, 0.02))) +
  theme(panel.grid = element_blank(),
        legend.position = c(0.85, 0.6)) +
  labs(
    subtitle = glue("Percent incomplete records by year",
                    "\nMissing begin or end coordinates"),
    x = NULL,
    tag = "A"
  )

x_ann <- 1975
y_ann <- 0.95 * max(dta_for_datacheck$length_mi)

p2 <- dta_for_datacheck |>
  ggplot() +
  geom_jitter(aes(year, length_mi),
              size = 0.5, alpha = 0.6) +
  annotate("text", x = x_ann, y = y_ann, 
           label = glue("Excludes {n_zero} tornadoes", 
           "\nmissing coordinate data",
           "\nin the years before 2005"),
           hjust = 0, size = 3, fontface = 'italic') +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  theme(panel.grid = element_blank(),
        legend.position = c(0.7, 0.2)) +
  labs(
    subtitle = "Track length by year: Coordinate data for more\nshort tracks have been detected in recent years",
    x = NULL,
    tag = "B"
  )

p3 <- dta_non_mapping |>
  ggplot() +
  geom_jitter(aes(year, tor_f_scale),
              size = 0.5, alpha = 0.6) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  theme(panel.grid = element_blank(),
        ) +
  labs(
    subtitle = "Force by year: more weak tornadoes\nhave been detected in recent years",
    x = NULL,
    tag = "C"
  )

p1 + p2 + p3 + 
  plot_annotation(
    title = "Changes in tornado detection over time in NC",
    caption = my_caption,
    theme = theme(plot.title = element_text(size = rel(2.0), face = "bold"))
  )

```

<br>

## Track length

The differences in recorded and calculated track length have changed over time: a higher proportion of the differences are very small after about 2007 (with the exception of the 100% artifacts below).

Some observations arising from @fig-compare-reported-and-calc-length-over-time:

* Panel A: from about 1994 to 2007 it seems NWS staff commonly (but not exclusively) used round numbers when recording track length: $[0.1, 0.2, 0.3, 0.4, 0.5, 1.0, 2.0, 3.0]$
* Panel B: records with non-zero recorded length but zero calculated length are plotted as 100% different. Why does this occur mainly between 1993 and 2007? Perhaps the NWE forecasting offices gained instruments that could better detect short tracks but not with great accuracy. Or perhaps they stopped sending people out to manually measure every event, instead relying on weather radars that did not allow sufficient accuracy in determining beginning and ending coordinates.
* Panel C: If the recorded track length matched the length I calculated using the coordinates were equal, all points would lay on the line.

See also @fig-track-length-ecdf NC tornado segment track measurement differences.

```{r, fig.height=8, fig.width=12}
#| label: fig-compare-reported-and-calc-length-over-time
#| fig-cap: "Comparing reported and calculated tornado track length over time"
#| fig-height: 8
#| fig-width: 12
#| column: page-inset-right

n_5mi <- dta_for_datacheck |>
  filter(length_diff > 5) |>
  nrow()

n_double <- dta_for_datacheck |>
  filter(pct_diff > 2) |>
  nrow()

n_calc_zero <- dta_for_datacheck |>
  filter(tor_length > 0 & length_mi < 1e-6) |>
  nrow()
  
p1 <- dta_for_datacheck |>
  filter(length_diff <= 5) |>
  ggplot() +
  geom_jitter(aes(year, length_diff),
              size = 0.5, alpha = 0.6) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  theme(panel.grid = element_blank()) +
  labs(
    subtitle = glue("Excluding {n_5mi} events w/ calculated and reported\nlength differing by more than 5 mi"),
    x = NULL,
    y = "Difference in miles",
    tag = "A"
  )

p2 <- dta_for_datacheck |>
  filter(pct_diff < 2) |>
  ggplot() +
  geom_jitter(aes(year, pct_diff),
              size = 0.5, alpha = 0.6) +
  scale_y_continuous(labels = label_percent(),
                     expand = expansion(mult = c(0, 0.02))) +
  theme(panel.grid = element_blank()) +
  labs(
    subtitle = glue("Excluding {n_double} events w/ calculated and reported\nlength differing by more than 200%",
                    "\n{n_calc_zero} events have non-zero recorded length but\nzero calculated length (shown as 100% diff)"),
    x = NULL,
    y = "Percent difference",
    tag = "B"
  )

p3 <- dta_for_datacheck |>
  filter(tor_length > 0,
         length_mi > 0) |>
  ggplot() +
  geom_jitter(aes(tor_length, length_mi, color = decade),
              size = 0.5, alpha = 0.6) +
  scale_x_continuous(trans = "log2",
                     expand = expansion(mult = c(0.01, 0.02))) +
  scale_y_continuous(trans = "log2",
                     labels = label_number(accuracy = 0.01)
                     #expand = expansion(mult = c(0, 0.02))
                     ) +
  theme(panel.grid = element_blank(),
        legend.position = c(0.8, 0.3)) +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  
  labs(
    subtitle = glue("Calcuated length_mi by recorded tor_length (mi)",
                    "\nExcluding {n_zero} observations with zero length"),
    x = "recorded length (log2 scale)",
    y = "calculated length (log2 scale)",
    color = NULL,
    tag = "C"
  )
  
p1 + p2 + p3 +
  plot_annotation(
    title = "How closely do reported and calculated track length agree over time?",
    subtitle = "The differences are smaller in recent years",
    caption = my_caption,
    theme = theme(plot.title = element_text(size = rel(2.0), face = "bold"))
  )

```

<br>

## Track length error

The improvement in measurement is visible in the reduction of error. @fig-compare-rse plots root square error (RSE)^["Root square error" is abs(error). I use this term, because I also summarize error by decade as "root *mean* square error" (RMSE)] on the X axis. The improvements are visible beginning in the 2000s both visually and in RMSE.

```{r, fig.height=8, fig.width=12}
#| label: fig-compare-rse
#| fig-cap: "Comparing track length error by decade: root square error as percent of tor_length (mi) by calculated length_mi"
#| fig-height: 8
#| fig-width: 12
#| column: page-inset-right

data_for_plot <- dta_for_datacheck |>
  filter(tor_length > 0,
         length_mi > 0) |>
  mutate(rse = sqrt((tor_length - length_mi)^2),
         rse_pct_of_tor_length = rse / tor_length) 

rmse_data <- data_for_plot |>
  reframe(rmse = sqrt(mean((tor_length - length_mi)^2)),
          x = min(rmse),
          y = max(rse_pct_of_tor_length),
            .by = decade)

data_for_plot |>
  ggplot() +
  geom_point(aes(rse, rse_pct_of_tor_length, color = decade),
              size = 0.5, alpha = 0.6,
             show.legend = FALSE) +
  geom_text(data = rmse_data,
            aes(x = 0.00005, y = 16, label = paste0("RMSE: ", round(rmse, digits = 2)), # min(x), y = max(y)
                hjust = 0),
            size = 3) +
  scale_x_continuous(trans = "log2",
                     #expand = expansion(mult = c(0.02, 0.02)),
                     labels = label_number(accuracy = 0.0001)) +
  scale_y_continuous(trans = "log2",
                     #expand = expansion(mult = c(0.02, 0.02)),
                     labels = label_number(accuracy = 0.0001)
                     ) +
  facet_wrap(~decade) +
  theme(panel.grid = element_blank(),
        plot.title = element_text(size = rel(2.0), face = "bold"),
        legend.position = c(0.8, 0.3)) +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  labs(
    title = "A reduction in error over time: RSE as percent of tor_length (mi)\nby RSE of calculated length_mi",
    subtitle = glue("Excluding {n_zero} observations with zero length"),
    x = "rse (log2 scale)",
    y = "rse_pct_of_tor_length (log2 scale)",
    color = NULL
  )

```

<br>
